A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Electron Beam Dynamics

### 2018-09-29 paper/prab.tex
first round of grammer edits

\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{todonotes}
\usepackage{siunitx}
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage{fancyvrb}
\usepackage{booktabs}
%\usepackage{cite}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{url}
%\usepackage{footmisc}
%\usepackage{multicol}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,snakes,backgrounds}
\usetikzlibrary{mindmap,trees}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{plotmarks}
\usepackage{listings}
\newcounter{lstmain}
\setcounter{lstmain}{1}

@@

\begin{figure}%[h!]
%\begin{center}%\normalsize
	\scalebox{0.8}{
	\pgfdeclarelayer{background}
	\pgfsetlayers{background,main}
	\begin{tikzpicture}[text=black]%[scale=0.05]%, transform shape]
	\input{figures/opt-framework_code.tex}
	\end{tikzpicture}
}
%\end{center}
\caption{Multi-objective framework: the pilot (master) solves the
	optimization problem specified in the input file by coordinating optimizer
	algorithms and workers running forward solves.}
\end{figure}\label{fig:framenetwork}

We developed a modular multi-objective software framework (see
  Fig.~\ref{fig:framenetwork}) where the core functionality is decoupled from
  the ``forward solver'' and the optimizer.
To that end, we use a master/slave mechanism where a master process governs a
  set of slave processes given some computational tasks to complete.
This separation allows to easily interchange optimizer algorithms, forward
  solvers and optimization problems.
A ``pilot'' coordinates all efforts between the optimization algorithm and the
  forward solver.
This forms a robust and general framework for massively parallel
  multi-objective optimization.
Currently the framework offers one concrete optimization algorithm, an
  evolutionary algorithm employing a \textsc{NSGAII} selector \cite{pisa}.
Normally, simulation based approaches are plagued by the trade-of between
  level of detail and time to solution.
We address this problem by using forward solvers with different time and
  resolution complexity.

The framework discussed here, incorporates the following three contributions:
%
\begin{enumerate}
  \item Implementation of a scalable optimization algorithm capable of
        approximating Pareto fronts in high dimensional spaces,
  \item design and implementation of a modular framework that is simple to use
        and deploy on large scale computation resources, and
  \item demonstrating the usefulness of the proposed framework on a real world
        application in the domain of particle accelerators.
\end{enumerate}

Here, we will focus on a real world application, the a high charge photoinjector 
  at the Argonne Wakefield Accelerator (AWA).

  %Both of which have an immediate and a
  %long-range impact, by optimizing important parameters for rf photoinjectors.

The next section introduces the notation of multi-objective optimization
  theory and describes the first implemented optimizer.
In Section~\ref{sec:framework} we discuss the implementation of the framework.
We introduce the employed forward-solver in Section~\ref{sec:forward-solver}.
A validation and a proof of concept application in the two beam dynamics problems 
mentioned above is discussed in Section~\ref{sec:experiments}.

@@

Optimization problems deal with finding one or more feasible solutions
  corresponding to extreme values of objectives.
If more than one objective is present in the optimization problem we call this
  a multi-objective optimization problem (MOOP).
A MOOP is defined as
%
\begin{align}
  \text{ min} \quad & \quad f_m({\bf x}), ~& m &= 1, \dots, M, \label{eq:moop:obj}\\
  \text{s.t.} \quad & \quad g_j({\bf x}) \geq 0, & j &= 0, \dots, J,
  \label{eq:moop:constr}\\
  \quad & \quad  x_i^L \leq {\bf x}=x_i \leq x_i^U,& i &=0, \dots, n
  \label{eq:moop:dvar} \text{,}
\end{align}
%
where $\bf f$ denotes the objectives (\ref{eq:moop:obj}),
  $\bf g$ the constraints (\ref{eq:moop:constr}),
  and $\bf x$ the design variables (\ref{eq:moop:dvar}).

Often, we encounter conflicting objectives complicating the concept of
  optimality.
To illustrate this, let us consider the problem of buying a car.
Naturally, we try to bargain the lowest price for the best performance,
  e.g.\ maximal horsepower or minimal fuel consumption.
This can be formulated as MOOP (\ref{eq:car-moop}).
%
\begin{equation}
  \begin{array}{cl}
  \min          & \quad \left[ \text{price}, -\text{performance} \right]^T \\
    \text{s.t.} & \quad \text{low}_\text{pr} \leq \text{price} \leq \text{high}_\text{pr}\\
                & \quad \text{low}_\text{pe} \leq \text{performance} \leq \text{high}_\text{pe}
  \end{array}
  \label{eq:car-moop}
\end{equation}

@@

Once the shape of the Pareto front has been determined the buyer can
  \textit{specify preference}, balancing the features by observing the
  effect on the optimality criteria, converging to the preferred solution.
This is called \textit{a posteriori} preference specification since we select
  a solution after all possible trade-offs have been presented to us.
An alternative is to specify preference \textit{a priori}, e.g., by
  weighting (specifying preference before solving the problem) and combining
  all objectives into one and applying a single-objective method to solve the
  problem (yielding only one solution).
In many situations preference is not known \textit{a priori} and an
  \textit{a posteriori} preference specification helps conveying a deeper
  understanding of the solution space.
The Pareto front can be explored and the impact of a trade-off decision then
  becomes visible.

@@

Since there already exist plenty of implementations of evolutionary algorithms,
  we decided to incorporate the PISA library \cite{pisa} into our
  framework.
One of the advantages of PISA is that it separates variator from selector,
  rendering the library expandable and configurable.
Implementing a variator was enough to use PISA in our framework and
  retain access to all available PISA selectors.
As shown in Fig.~\ref{fig:varsel}, the selector is in charge of ordering a
  set of $d$-dimensional vectors and selecting the $k$ fittest individuals
  currently in the population.
The performance of a selector depends on the number of objectives and the
  surface of the search space.
So far, we used an NSGA-II selector \cite{dpam:02} exhibiting satisfactory
  convergence performance.

@@

Simulation based multi-objective optimization problems are omnipresent in
  research and the industry.
The simulation and optimization problems arising in such problems are in
  general very big and computationally demanding.
This motivated us to design a massively parallel general purpose framework.
The key traits of such a design can be summarized as:
%
\begin{itemize}
  \item support any multi-objective optimization method,
  \item support any function evaluator: simulation code or measurements,
  \item offer a general description/specification of objectives, constraints
        and design variables,
  \item run efficiently in parallel on current large-scale high-end clusters
        and supercomputers.
\end{itemize}
%

@@

%
The \textsc{Pilot} component acts as a bridge between the optimizer and
  forward solvers, providing the necessary functionality to handle passing
  requests and results between the \textsc{Optimizer} and the
  \textsc{Simulation} modules.

We implemented the framework in \texttt{C++}, utilizing features like template
  parameters to specify the composition of the framework (shown in Listing 1).

@@

typedef Pilot< Input_t, Opt_t, Sim_t > Pilot_t;
\end{code}
%\end{minipage}

The framework provides ``default'' implementations that can be controlled via
  command line options.
Due to its modular design, all components can be completely customized.

@@

After a process gets appointed a role it starts a polling loop asynchronously
  checking for appropriate incoming requests.
To that end a \textsc{Poller} interface helper class has been introduced.
The \textsc{Poller} interface consists of an infinite loop that checks
  periodically for new MPI messages.
Upon reception a new message is immediately forwarded to the appropriate
  handler, the \texttt{onMessage()} method.
The method is called with the \texttt{MPI\_Status} of the received message and
  a \texttt{size\_t} value specifying different values depending on the value
  of the \texttt{MPI\_Tag}.
The \textsc{Poller} interface allows the implementation of special methods
  (denoted \textit{hooks}) determining the behavior of the polling process,
  e.g.\ for actions that need to be taken after a message has been handled.
Every \textsc{Poller} terminates the loop upon receiving a special MPI tag.

@@

In most cases \textsc{Simulation} implementations are simple wrappers to run
  an existing ``external'' simulation code using a set of design variables as
  input.
As for the \textsc{Optimizer} component there exists a base class, labeled
  \texttt{Simulation} as common basis for all \textsc{Simulation}
  implementations.
In addition, this component also inherits from the \texttt{Worker} class,
  already implementing the polling protocol for default worker types.
As shown in the API in Listing 3 the \texttt{Worker} class expects an
  implementation to provide implementations for those three methods.

@@

First, upon receiving a new job, the \texttt{Worker} will call the
  \texttt{run} method on the \textsc{Simulation} implementation.
This expects the \textsc{Simulation} implementation to run the simulation in a
  \textit{blocking} fashion, meaning the method call blocks and does not return
  until the simulation has terminated.
Subsequently, the \texttt{Worker} calls \texttt{collectResults}, where the
  \textsc{Simulation} prepares the result data, e.g., parsing output files,
  and stores the requested information in a \texttt{reqVarContainer\_t} data
  structure.
Finally, the results obtained with \texttt{getResults} are sent to the
  \textsc{Pilot}.
As before, the serialized data is exchanged using MPI point-to-point
  communication using a specific set of communicators.

@@

We aimed at an easy and expressive way for users to specify multi-objective
  optimization problems.
Following the principle of keeping metadata (optimization and simulation
  input data) together, we decided to embed the optimization problem
  specification in the simulation input file by prefixing it with special
  characters, e.g., as annotations prefixed with a special character.
In some cases it might not be possible to annotate the simulation input file.
By providing an extra input file parser, optimization problems can be read
  from stand-alone files.

To allow arbitrary constraints and objective expressions, such as
%
%{0.2cm}
\begin{Verbatim}[fontsize=\scriptsize]
  name: OBJECTIVE,
        EXPR="5 * average(42.0, "measurement.dat") + ENERGY";
\end{Verbatim}
%\vspace{0.2cm}
%
\noindent
  we implemented an expression parser using Boost
%  Spirit\footnote{\url{http://boost-spirit.com/}}.
In addition to the parser, we need an evaluator able to evaluate an expression,
  given a parse tree and variable assignments to an actual value.
Expressions arising in multi-objective optimization problems usually evaluate
  to booleans or floating point values.
The parse tree, also denoted abstract syntax tree (AST), is constructed
  recursively while an expression is parsed.
Upon evaluation, all unknown variables are replaced with values, either
  obtained from simulation results or provided by other subtrees in the AST.
In this stage, the AST can be evaluated bottom-up and the desired result is
  returned after processing the root of the tree.

>>>>

\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{todonotes}
\usepackage{siunitx}
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage{fancyvrb}
\usepackage{booktabs}
%\usepackage{cite}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{url}
%\usepackage{footmisc}
%\usepackage{multicol}
%\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,snakes,backgrounds}
\usetikzlibrary{mindmap,trees}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{plotmarks}
\usepackage{listings}
\newcounter{lstmain}
\setcounter{lstmain}{1}

@@

\begin{figure}%[h!]
%\begin{center}%\normalsize
	\scalebox{0.8}{
	\pgfdeclarelayer{background}
	\pgfsetlayers{background,main}
	\begin{tikzpicture}[text=black]%[scale=0.05]%, transform shape]
	\input{figures/opt-framework_code.tex}
	\end{tikzpicture}
}
%\end{center}
\caption{Multi-objective framework: the pilot (master) solves the
	optimization problem specified in the input file by coordinating optimizer
	algorithms and workers running forward solves.}
\label{fig:framenetwork}
\end{figure}

A modular multi-objective software framework was developed (see
  Fig.~\ref{fig:framenetwork}) where the core functionality is decoupled from
  the ``forward solver'' and the optimizer.
To that end, we use a master/slave mechanism where a master process governs a
  set of slave processes given some computational tasks to complete.
This separation allows easy interchange of optimization algorithms, forward
  solvers and optimization problems.
A ``pilot'' coordinates all efforts between the optimization algorithm and the
  forward solver.
This forms a robust and general framework for massively parallel
  multi-objective optimization.
Currently the framework offers one concrete optimization algorithm, an
  evolutionary algorithm employing a \textsc{NSGA-II} selector \cite{pisa}.
Normally, simulation based approaches are plagued by the trade-of between
  level of detail and time to solution.
We address this problem by using forward solvers with different time and
  resolution complexity.

The framework discussed here, incorporates the following three contributions:
%
\begin{enumerate}
  \item Implementation of a scalable optimization algorithm capable of
        approximating Pareto fronts in high dimensional spaces,
  \item design and implementation of a modular framework that is simple to use
        and deploy on large scale computational resources, and
  \item demonstrating the usefulness of the proposed framework on a real world
        application in the domain of particle accelerators,
        with the optimization problem set as 
        the high charge photoinjector at the Argonne Wakefield Accelerator (AWA). 
\end{enumerate}

The next section introduces the notation of multi-objective optimization
theory and describes the first implemented optimizer.
In Section~\ref{sec:framework} the implementation of the framework is discussed.
We introduce the employed forward-solver in Section~\ref{sec:forward-solver}.
A validation and a proof of concept application in the two beam dynamics problems 
mentioned above is discussed in Section~\ref{sec:experiments}.

@@

Optimization problems deal with finding one or more feasible solutions
  corresponding to extreme values of objectives.
If more than one objective is present in the optimization problem this is called
  a multi-objective optimization problem (MOOP).
A MOOP is defined as
%
\begin{align}
  \text{ min} \quad & \quad f_m({\bf x}), ~& m &= 1, \dots, M, \label{eq:moop:obj}\\
  \text{s.t.} \quad & \quad g_j({\bf x}) \geq 0, & j &= 0, \dots, J,
  \label{eq:moop:constr}\\
  \quad & \quad  x_i^L \leq {\bf x}=x_i \leq x_i^U,& i &=0, \dots, n
  \label{eq:moop:dvar} \text{,}
\end{align}
%
where $\bf f$ denotes the objectives (\ref{eq:moop:obj}),
  $\bf g$ the constraints (\ref{eq:moop:constr}),
  and $\bf x$ the design variables (\ref{eq:moop:dvar}).

Often, conflicting objectives are encountered, and this complicates the concept of
  optimality.
To illustrate this, let us consider the problem of buying a car.
Naturally, we try to bargain the lowest price for the best performance,
  e.g.\ maximal horsepower or minimal fuel consumption.
This can be formulated as MOOP (\ref{eq:car-moop}).
%
\begin{equation}
  \begin{array}{cl}
  \min          & \quad \left[ \text{price}, -\text{performance} \right]^T \\
    \text{s.t.} & \quad \text{low}_\text{pr} \leq \text{price} \leq \text{high}_\text{pr}\\
                & \quad \text{low}_\text{pe} \leq \text{performance} \leq \text{high}_\text{pe}
  \end{array}
  \label{eq:car-moop}
\end{equation}

@@

Once the shape of the Pareto front has been determined, the buyer can
  \textit{specify preference}, balancing the features by observing the
  effect on the optimality criteria, converging to the preferred solution.
This is called \textit{a posteriori} preference specification since we select
  a solution after all possible trade-offs have been presented to us.
An alternative is to specify preference \textit{a priori}, e.g., by
  weighting (specifying preference before solving the problem) and combining
  all objectives into one and applying a single-objective method to solve the
  problem (yielding only one solution).
In many situations preference is not known \textit{a priori} and an
  \textit{a posteriori} preference specification helps conveying a deeper
  understanding of the solution space.
The Pareto front can be explored and the impact of a trade-off decision then
  becomes visible.

@@

Since there already exist plenty of implementations of evolutionary algorithms,
  we decided to incorporate the PISA library \cite{pisa} into our
  framework.
One of the advantages of PISA is that it separates variator from selector,
  rendering the library expandable and configurable.
Implementing a variator was enough to use PISA in our framework and
  retain access to all available PISA selectors.
As shown in Fig.~\ref{fig:varsel}, the selector is in charge of ordering a
  set of $d$-dimensional vectors and selecting the $k$ fittest individuals
  currently in the population.
The performance of a selector depends on the number of objectives and the
  surface of the search space.
So far, the NSGA-II selector \cite{dpam:02} has been used and exhibits satisfactory
  convergence performance.

@@

Simulation based multi-objective optimization problems are omnipresent in
  research and industry.
The simulation and optimization problems arising in such problems are in
  general very big and computationally demanding.
This motivated us to design a massively parallel general purpose framework.
The key traits of such a design can be summarized as:
%
\begin{itemize}
  \item support any multi-objective optimization method,
  \item support any function evaluator: simulation code or measurements,
  \item offer a general description/specification of objectives, constraints
        and design variables,
  \item run efficiently in parallel on current large-scale high-end clusters
        and supercomputers.
\end{itemize}
%

@@

%
The \textsc{Pilot} component acts as a bridge between the optimizer and
  forward solvers, providing the necessary functionality to handle passing
  requests and results between the \textsc{Optimizer} and the
  \textsc{Simulation} modules.
The framework was implemented in \texttt{C++}, utilizing features like template
  parameters to specify the composition of the framework (shown in Listing 1).

@@

typedef Pilot< Input_t, Opt_t, Sim_t > Pilot_t;
\end{code}
%\end{minipage}
The framework provides ``default'' implementations that can be controlled via
  command line options.
Due to its modular design, all components can be completely customized.

@@

After a process gets appointed a role it starts a polling loop asynchronously
  checking for appropriate incoming requests.
To that end a \textsc{Poller} interface helper class has been introduced.
The \textsc{Poller} interface consists of an infinite loop that checks
  periodically for new MPI messages.
Upon reception a new message is immediately forwarded to the appropriate
  handler: the \texttt{onMessage()} method.
The method is called with the \texttt{MPI\_Status} of the received message and
  a \texttt{size\_t} value specifying different values depending on the value
  of the \texttt{MPI\_Tag}.
The \textsc{Poller} interface allows the implementation of special methods
  (denoted \textit{hooks}) determining the behavior of the polling process,
  e.g.\ for actions that need to be taken after a message has been handled.
Every \textsc{Poller} terminates the loop upon receiving a special MPI tag.

@@

In most cases \textsc{Simulation} implementations are simple wrappers to run
  an existing ``external'' simulation code using a set of design variables as
  input.
As for the \textsc{Optimizer} component there exists a base class, labeled
  \texttt{Simulation} as common basis for all \textsc{Simulation}
  implementations.
In addition, this component also inherits from the \texttt{Worker} class,
  already implementing the polling protocol for default worker types.
As shown in the API in Listing 3, the \texttt{Worker} class expects an
  implementation to provide implementations for those three methods.

@@

First, upon receiving a new job, the \texttt{Worker} will call the
  \texttt{run} method on the \textsc{Simulation} implementation.
This expects the \textsc{Simulation} implementation to run the simulation in a
  \textit{blocking} fashion, meaning the method call blocks and does not return
  until the simulation has terminated.
Subsequently, the \texttt{Worker} calls \texttt{collectResults}, where the
  \textsc{Simulation} prepares the result data, e.g. parsing output files,
  and stores the requested information in a \texttt{reqVarContainer\_t} data
  structure.
Finally, the results obtained with \texttt{getResults} are sent to the
  \textsc{Pilot}.
As before, the serialized data is exchanged using MPI point-to-point
  communication using a specific set of communicators.

@@

We aimed at an easy and expressive way for users to specify multi-objective
  optimization problems.
Following the principle of keeping metadata (optimization and simulation
  input data) together, we decided to embed the optimization problem
  specification in the simulation input file by prefixing it with special
  characters, e.g. as annotations prefixed with a special character.
In some cases it might not be possible to annotate the simulation input file.
By providing an extra input file parser, optimization problems can be read
  from stand-alone files.

To allow arbitrary constraints and objective expressions, such as
%
%{0.2cm}
\begin{Verbatim}[fontsize=\scriptsize]
  name: OBJECTIVE,
        EXPR="5 * average(42.0, "measurement.dat") + ENERGY";
\end{Verbatim}
%\vspace{0.2cm}
%
\noindent
  we implemented an expression parser using Boost Spirit\footnotemark \footnotetext{test}
  	%http://boost-spirit.com/}.
In addition to the parser, we need an evaluator able to evaluate an expression,
  given a parse tree and variable assignments to an actual value.
Expressions arising in multi-objective optimization problems usually evaluate
  to booleans or floating point values.
The parse tree, also denoted abstract syntax tree (AST), is constructed
  recursively while an expression is parsed.
Upon evaluation, all unknown variables are replaced with values, either
  obtained from simulation results or provided by other subtrees in the AST.
In this stage, the AST can be evaluated bottom-up and the desired result is
  returned after processing the root of the tree.
